Fairness in Machine Learning SS21

### 2021-05-09 references/vanilla_presentation .Rmd
presentation slides

---
title: "Fairness in Machine Learning WS20/21"
subtitle: "Fostering Notions of Dynamic Fairness via Structural Causal Models"
date: "03/05/2021"
author: "Daniel Saggau"
output:
  beamer_presentation:
    theme: Madrid
    colortheme: spruce
---

@@

@@

- Fairness in dynamical system has become important because equalizing true positive rate at each step does not converge as fast in systems with e.g. population dynamics 
- Research on dynamical systems has focused on markov decision processes
- Prior research on causal fairness has focused in static systems 
- Recently Creager et al. (2020) proposed a structural causal model for dynamical systems

@@

- Fair-MDP
- Judea Pearl:do-calculus
-

# SCM vs Causal Bayesian Network (Probabilistic Model)

@@

# Time and Causal Inference

- Structural causal models are abstractions of physical processes
-

@@

- Cyclic Structural Causal Models with actual reinforcing loops
- Semi-Deterministic SCMs (deterministic -> all variables are known)

>>>>

---
title: "Fairness in Machine Learning SS21"
subtitle: "Beyond Static Fairness"
date: "03/05/2021"
author: "Daniel Saggau"
output:
  beamer_presentation:
    theme: Madrid
    colortheme: spruce
---

@@

# Dynamical Repercussions

- Apple card discriminating women

@@

- Fairness in dynamical system has become important because equalizing true positive rate at each step does not converge as fast in systems with e.g. population dynamics 
- Research on dynamical systems has focused on markov decision processes
- Prior research on causal fairness has focused in static systems

# Fairness-Gym

- Environment to simulate fairness

@@

- Fair-MDP
- Judea Pearl:do-calculus

# Graphical Illustration

```{tikz,fig.align="center",fig.cap="Probabilistic Model", echo =F}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
\begin{tikzpicture}[
    sharp corners=2pt,
    inner sep=7pt,
    node distance=3cm,
    >=latex]
\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
\node[my node] (A){A};
\node[my node,right =1 cm of A](C){C};
\node[my node] at ($(A)!0.5!(C)-(0pt,1.5cm)$) (T) {T};
\draw[->] (A) -- (T);
\draw[->] (A) -- (C);
\draw[->] (T) -- (C);
\end{tikzpicture}
```

```{tikz,fig.cap="Structural Causal Model", fig.align="center", echo =F}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
\usetikzlibrary{calc}
\usetikzlibrary{positioning}
\usetikzlibrary{arrows}
  \begin{tikzpicture}[
    sharp corners=2pt,
    inner sep=6pt,
    node distance=2cm,
    >=latex]
\tikzstyle{my node}=[draw, shape = circle, minimum height=1cm,minimum width=1cm]
\tikzstyle{latent}=[draw, shape = rectangle, minimum height=1cm,minimum width=1cm]
\node[my node, fill=gray!30] (A){A};
\node[latent,left of=A](UA){$U_A$};
\node[my node, fill=gray!30] at ($(A)!0.5!(A)-(0pt,1.5cm)$) (T) {T};
\node[my node,right = 1 cm of T, fill=gray!30](C){C};
\node[latent,left = 1cm of T](UT){$U_T$};
\node[latent,right = 1cm of A](UC){$U_C$};
\draw[->] (A) -- (T);
\draw[->] (UA) -- (A);
\draw[->] (UT) -- (T);
\draw[->] (UC) -- (C);
\draw[->] (A) -- (C);
\draw[->] (T) -- (C);
\end{tikzpicture}
```

# Interventions

- Atomic Intervention 
- Policy Intervention 
- Off-Policy Intervention (model-based; model-free)

# Results Off Policy Intervention - Lambda

* Introduce trade off parameter lambda 
* V of Pi is the overall objective, Pi is our policy and U is utility

$$V_\pi =U - \lambda\delta_{EQOPP}$$

# Causal Hierachy

 |Method          | Action |  Example | Usage | 
|------------------|-------------|--------------------|-------------------|
| Association $P(a|b)$               | Co-occurrence             | What happened...               |(Un-)Supervised ML, BN, Reg.  
| Intervention $P(a|do(b),c)$       | Do-manipulation           | What happens if ...            |CBN,MDP,RL    
| Counterfactual $P(a_b|a`,b`)$     | Hypotheticals   | What would have happened if...           | SCM ,PO

Table: Pearls Hierachy of Causation (2009)

@@

@@

- Cyclic Structural Causal Models with actual reinforcing loops
- Semi-Deterministic SCMs (deterministic -> all variables are known)
-
